%!TEX root = ../username.tex
\chapter{A Markov Model for Melody Generation} \label{markov}

\section{Setup of the Model} \label{markov:setup}

In this model, we use two Markov processes: one to determine interval changes between notes to generate pitches, and the other to generate rhythms.
To create the interval transition matrix, we iterate over the source -- in our case a corpus of music by J.S. Bach -- keeping track of the previous interval at each step and incrementing the matrix at position $a_{i,j}$, where $i$ is the previous interval and $j$ is the current interval between notes.
Note that for a matrix $m$, $m_{i,j}$ indicates the element in row $i$ and column $j$.
The rhythmic transition matrix is similarly generated.
While iterating over the source to generate the interval transition matrix, we can also build the rhythmic transition matrix.
To do this we keep track of the duration of the previous rhythm at each step and increment the transition matrix at position $b_{k,l}$ where $k$ is the previous rhythmic duration and $l$ is the current rhythmic duration.

Because the number of possible intervals and rhythms are theoretically infinite and the transition matrix would be very sparse, we do not want to try to store a matrix capable of holding every possible transition.
Instead, we can use a hash table to only store transitions that actually appear in the source.

An example rhythmic transition matrix is given in Figure \ref{fig:rhythmTransitionMatrix}, based on the \textit{Little Fugue in G Minor} by J.S. Bach.

\begin{figure}
	\centering
	\begin{tabular}{c | c c c c c c c c}
		& 0.25 & 0.5 & 0.75 & 1.0 & 1.25 & 1.5 & 2.0 & 2.25\\
		\hline
		0.25 & 606 & 21 & 1 & 2 & 1 & 2 & 0 & 1\\
		0.5 & 23 & 109 & 0 & 7 & 1 & 0 & 0 & 0\\
		0.75 & 1 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
		1.0 & 1 & 6 & 0 & 3 & 0 & 3 & 2 & 0\\
		1.25 & 2 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
		1.5 & 0 & 5 & 0 & 0 & 0 & 0 & 0 & 0\\
		2.0 & 0 & 0 & 0 & 2 & 0 & 0 & 0 & 0\\
		2.25 & 1 & 0 & 0 & 0 & 0 & 0 & 0 & 0
	\end{tabular}
	\caption{An example rhythmic transition matrix. A value of $1.0$ indicates a quarter note.}
	\label{fig:rhythmTransitionMatrix}
\end{figure}

\section{Generation} \label{markov:generation}

After the transition matrices are created, the program randomly selects a rhythmic value, based on the probabilities that correspond with the previous note, and a pitch.
The pitch value comes from the interval from the previous note.
The interval is chosen based on the probabilities that correspond with the previous interval.

The program stops generating notes when the stopping criteria are met: the generated melody is at least eight measures, the melody ends at the end of a measure, and the last note is a tonic chord tone.
Because the generated melody could theoretically be infinite, the program terminates generation after one-hundred measures are generated whether or not the stopping criteria are met.
